\bf Adapting the Chumbley Score to match striae on Land Engraved Areas (LEAs) of bullets

### 2018-09-17 jfs_submission/paper-submission-original/paper-title.tex
re-arranging the files

% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
%\usepackage[numbers]{natbib}
\usepackage[nomarkers]{endfloat}
\usepackage{natbib}
\setcitestyle{numbers} 
\makeatletter % Reference list option change
\renewcommand\@biblabel[1]{#1. } % from [1] to 1
\makeatother %

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{adjustbox}
\usepackage{multirow}
\usepackage{subfig}
\usepackage[table,xcdraw]{xcolor}
\usepackage{wrapfig}
\usepackage{float}
%\usepackage{colortbl}
\usepackage[colorlinks]{hyperref}
\hypersetup{
  colorlinks=true,
  citecolor=black,
  linkcolor=black,
  urlcolor=blue}

\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{url} % not crucial - just used below for the URL

\usepackage{etoolbox}% http://ctan.org/pkg/etoolbox
\makeatletter
\patchcmd{\subsection}{\bfseries}{\relax}{}{}% Non-bold \subsection
\patchcmd{\subsubsection}{\bfseries}{\relax}{}{}% Non-bold \subsection
\makeatother

@@

\if0\blind
{
  \title{\bf Adapting the Chumbley Score to match striae on Land Engraved Areas of
bullets}

  \author{
        Ganesh Krishnan,\\ %\\
    Master of Science (M.Sc.)\\
    % \\
     and \\     Prof.~Heike Hofmann,\\ %\\
    Doctor of Philosophy (Ph.D.)\\
    % \\
      }
    \maketitle
  \bigskip
  \begin{center}
  {
  \textbf{Affiliation(s)}\\
  Department of Statistics, Iowa State University and CSAFE (Center for
  Statistics and Applications in Forensic Evidence)\\
  \bigskip
  \textbf{Source of funding}\\
  The authors gratefully acknowledge funding from CSAFE through
  cooperative agreement \# 70NANB15H176 between NIST and Iowa State
  University.\\
  \bigskip
  \textbf{Acknowledgement}\\
  We would like to thank the efforts of CSAFE and Roy J Carver Microscopy
  lab in scanning the Hamby-44 set and providing the scans to us. We would
  also like to thank NIST for providing the Hamby-44 bullets to us for
  scanning. We would also like to thank the two anonymous reviewers for
  their helpful suggestions in improving the paper.\\
  \bigskip
  \textbf{Information on presentation of work}\\
  This work was presented in the form of a poster in the AFTE 2018 49th
  Annual Training Seminar, held from June 3rd to June 8th, 2018 in
  Charleston, West Virginia
  }
  \end{center}

@@

  % \bigskip

  % \bigskip
\newpage

\if0\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
  \title{\LARGE\bf Adapting the Chumbley Score to match striae on Land Engraved Areas of
bullets}
  \end{center}
  \bigskip
  \maketitle
} \fi

\bigskip
\begin{abstract}
The same-source problem remains a major challenge in forensic toolmark
and firearm examination. Technological advances in surface imaging allow
measurements of 3D surfaces at previously unforeseen resolutions and
enable digitized imaging. Here, we investigate the applicability of the
Chumbley scoring method \cite{hadler}, developed for screwdriver
markings, for same-source identification of bullet striae. We provide
methods to identify parameters that minimize the error rates for
matching of LEAs using Hamby datasets 44 and 252 measured by NIST and
CSAFE. We suggest a remedial algorithm to alleviate the problem of
failed tests in the method, increasing both the power of the test and
reducing error rates. Type II error rates of the proposed method improve
by more than one third (Type I error of 0.05) and are on average about
0.22. This puts the proposed method on similar footing as other single
feature matching approaches in the literature.
\end{abstract}

\noindent%
{\it Keywords:} forensic science, toolmark, cross-correlation, Mann-Whitney U statistic, land engraved areas (LEAs), algorithm, signatures, same-source problem
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!

\newcommand{\hh}[1]{{\textcolor{orange}{#1}}}
\newcommand{\gk}[1]{{\textcolor{blue}{#1}}}
\newcommand{\cited}[1]{{\textcolor{red}{#1}}}
\setlength\parindent{0pt}

\newpage

Same-source analyses are a major part of a Forensic Toolmark Examiner's
job. In current practice, examiners make these comparisons by visual
inspection under a comparison microscope and come to one of the
following four conclusions: identification, inconclusive, elimination or
unsuitable for examination \cite{afte-toolmarks1998}. These conclusions
are made on the basis of ``unique surface contours'' of the two
toolmarks being in ``sufficient agreement'' \cite{afte-toolmarks1998}.
AFTE describes the term ``sufficient agreement'' as the possibility of
another tool producing the markings under comparison, as practically
impossible \cite{afte-toolmarks1998}. Potential subject bias in the
assessment as well as the lack of specified error rates are the main
points of criticisms first raised by the National Research Council in
2009 \cite{NAS:2009} and later emphasized further by the President's
Council of Advisors on Science and Technology \cite{pcast2016}.
{\textcolor{blue}{need a sentence on striated marks}}

Technological advances, such as profilometers and confocal microscopy,
have made it possible to capture 3D surfaces in a high-resolution
digitized form. This technology has become more accessible over the last
decade, and has made its way into topological images of ballistics
evidence, such as bullet lands and breech faces
\cite{DeKinder1, DeKinder2, Bachrach1, vorburger2016}. Digitized images
of 3D surfaces form the basis of statistical analysis of toolmarks.
Statistical approaches based on this data remove both subjectivity from
the assessment and allow a quantification of error rates for false
positive and false negative identifications.

Methods for matching striated marks for a variety of tools have been
studied in the literature (see \autoref{tab:toolmarks-ER} for an
overview): in \cite{manytoolmarks1} and \cite{chumbley} digitized
screwdriver marks have been analyzed using a profilometer; in
\cite{manytoolmarks2} 3D marks from screwdriver, tongue and groove
pliers captured using a confocal microscope have been investigated;
digitized marks from slip-joint pliers generated by a surface
profilometer have been investigated in \cite{afte-chumbley}.

These digitized markings allow the use of statistical methods to
quantify the scientific mechanism of comparing markings and serve as
basis for an error rate calculation. In \cite{manytoolmarks2} a relative
distance metric is defined and used as similarity measure between two
toolmarks. This approach is expanded in \citet{manytoolmarks1}: a set of
small segments in the markings of two toolmarks are extracted and
similarity is compared using a maximum Pearson correlation coefficient.
The Chumbley scoring method, first introduced in \cite{chumbley}, uses a
similar but more extensive framework based on a Mann-Whitney U test of
the resulting correlation coefficients. This approach is
non-deterministic, because segments are chosen randomly. In
\cite{hadler} the score is fixed to be deterministic for each pair of
toolmarks by choosing segments for comparison systematically. This
approach also ensures independence between segments of striae.

\begin{table}[ht]
\centering
\resizebox{\textwidth}{!}{\begin{tabular}{rlrrr}
  \hline
Research paper & Method & Data Source & False Positives & False Negatives\\
  \hline
 \textbf{Faden et al. (2007)} \cite{manytoolmarks1} & Maximum Pearson & Screwdrivers & & \\ & Correlation &  & - & - \\ \hline
 \textbf{Chumbley et al. (2010)} \cite{chumbley} & Randomized & Screwdrivers & & \\ (Same-Surface Same-Angle) & Chumbley Score & & 2.3\% & 8.9\%  \\ \hline
  \textbf{Grieve et al. (2014)}
 \cite{afte-chumbley} & Randomized & Slip-joint  &  & \\ & Chumbley Score &  & - & - \\ \hline 
  \textbf{Hadler \& Morris (2017)} \cite{hadler} & Deterministic & Screwdrivers & & \\ (Same-Surface Same-Angle) & Chumbley Score & & 0\% & 6\%  \\ \hline
  \textbf{Bachrach et al. (2010)} \cite{manytoolmarks2} & Similarity Measure & Screwdrivers & & \\ (Different Surfaces-same angle) & Relative Distance Metric & & 5.9\% &  9.4\% \\ (Same Surfaces-same angle)&  & & 0.22\% &  0\% \\
   \hline
\end{tabular}}
\caption{\label{tab:toolmarks-ER} Error Rates in same-source analyses for  striated toolmarks reported in the literature. The top four reported papers use some variation of the Chumbley score method.}
\end{table}

In this paper, we are investigating the applicability of the Chumbley
scoring method by \citet{hadler} to assess striation marks on bullet
lands for same-source identification. Striation marks on bullets are
made by impurities in the barrel. As the bullet travels through the
barrel, these imperfections leave ``scratches'' on the bullet surface
(see top of \autoref{fig:rgl}). Typically, only striation marks in the
land engraved areas (LEAs) are considered \citep{afte-article1992}.
Bullet lands are depressed areas between the grooves made by the rifling
action of the barrel. Compared to toolmarks made by screwdrivers
striation marks on bullets are typically much smaller, both in length
and in width. Bullets also have a curved cross-sectional topography.

In same-source comparisons this curvature is usually removed using some
form of Gaussian filter \citep{ma2004} or non-parametric smoothing
\citep{aoas}. An overview of some of the error rates reported in the
literature on bullet matching is given in \autoref{tab:bullets-ER}.
\citet{chu2013} use an automatic method for counting consecutive
matching striae (CMS). The authors report an error rate of 52\% for
known same-source land comparisons to be (incorrectly) identified as
different-source (false negative) and zero false positives for known
different-source lands. \citet{ma2004} and \citet{vorburger2011} discuss
CCF (cross-correlation function) and its discriminating power and
applicability for same-source analyses of bullets, but do not provide
any error rates in their discussion. \citet{aoas} use multiple features,
such as CCF, CMS, D (distance measure), etc. in a random forest based
method and compare every land against every other land of digitized
versions of Hamby 252 and Hamby 44 \citep{hamby} published on the NIST
Ballistics Database \citep{nist}. The authors report an out-of-bag
overall error rate of 0.46\%, comprised of an error rate of 30.05\% of
same-source pairs that were not identified and an error rate of 0.026\%
of different-source pairs that were incorrectly identified as
same-source.

\begin{table}[ht]
\centering
\resizebox{\textwidth}{!}{\begin{tabular}{rlrrr}
  \hline
Method & Data Source & False Positives & False Negatives\\
  \hline
  \textbf{\citet{aoas}} & Hamby 252 (NIST scans) \\
Consecutive Matching Bullets Striae (CMS) &   &  6.25\% &  33.85\% \\ 
Consecutive Non-matching Striae (CNMS)  &  &  6.26\% & 35.42\%  \\ 
Average Distance (D) &   &   6.25\% & 45.83\% \\ 
Cross-correlation Function (CCF) & &  6.25\% & 17.71\%  \\ 
Sum of Peaks (S) &  &  6.25\% &  18.23\% \\
   \hline
   \textbf{\citet{chu2013}} & Hamby 252 (NIST scans) \\
Consecutive Matching Bullets Striae (CMS) &  &  0\% &  52\% \\ 
   \hline
   \textbf{\citet{chu2010}} & \small{6 types of firearms, 2 types of ammunition} \\
   \textbf{\citet{ma2004}} & \small{NIST standard bullet SRM 8240} \\
Cross-Correlation Function (CCF) &   &  - &  - \\ 
    \hline
\end{tabular}}
\caption{\label{tab:bullets-ER} Error Rates in same-source single feature land-to-land analysis reported in the literature. Note that only error rates from the same data source can be compared directly between methods. Error rates from different sources are influenced by type of firearm and ammunition used.}
\end{table}

The Chumbley score provides us with another approach in the same-source
assessment of bullet striation marks. \citet{chumbley} compare two
toolmarks for same-source. The data for this study was obtained from 50
sequentially manufactured screwdriver tips. \citet{chumbley} report
error rates for markings made by the tips at different angles. For
markings made at 30 degree the authors report an average false negative
error rate of 0.089 and an average false positive error rate of 0.023.
For marks made under angles of 60 and 85 degrees, respectively, the
false negatives error rate is 0.09 while the rate of false positives
decreases to 0.01. The paper by \citet{hadler} is based on the same data
but the authors focus on markings made under the same angle. The error
rates associated with the deterministic version of the score are
reported as 0.06 for false negatives and 0 for false positives.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{images/B6-B2-L6-rescaled.png}

\caption{\label{fig:rgl} Image of a bullet land from a confocal light microscope at 20 fold magnification (top) and a chart of the corresponding signature of the same land (bottom). The dotted lines connect some peaks visible in both visualizations.}

\end{figure}

In this paper we evaluate the adaptability of the Chumbley score as a
measure to quantify similarity in land engraved areas (LEAs) on bullets.
For that we briefly introduce the deterministic method suggested by
\citet{hadler} in section 2. In the process we provide methods to
identify parameters that minimize the error rates. We then investigate
persistent scenarios in which the method proposed by \citet{hadler}
fails to come to a result. We go on to provide a solution to the failed
tests problem, consequently increasing the power of the test and
reducing error rates in the process. We set up a testing framework to
compare the performance of the two algorithms in section 3 and finally
discuss results in section 4.

\hypertarget{methods}{%
\section*{Methods}\label{methods}}
\addcontentsline{toc}{section}{Methods}

\hypertarget{scans-for-land-engraved-areas}{%
\subsection*{\texorpdfstring{\emph{Scans for land engraved
areas}}{Scans for land engraved areas}}\label{scans-for-land-engraved-areas}}
\addcontentsline{toc}{subsection}{\emph{Scans for land engraved areas}}

Comparisons of striae from bullets are usually based on comparisons of
striae in land engraved areas, which are extracted in form of cross
sections, called \emph{profiles} \citep{aoas,ma2004}. Bullet striae are
most pronounced at the bottom of the bullet heel (because the heel has
the most contact with the insides of the barrel). Howver, these areas
are also affected by the break off due to friction effects between
barrel and the bullet. An optimal cross section is chosen orthogonally
to the striae, close to the heel while avoiding break off as shown in
\autoref{fig:rgl}, see also \citep{aoas} for mathematical details of the
extraction.

From profiles bullet \emph{signatures} \citep{chu2013,aoas} are
extracted as residuals of a loess fit or Gaussian filter. This
effectively removes topographic structure from the data in the attempt
to increase the signal to noise ratio. \autoref{fig:rgl} shows how the
signature from a bullet land (bottom) lines up with the image of the
land (top) from which it was extracted. We can see in the figure how the
depth and relative position of the striation markings seen in the image
are interpreted as peaks and valleys in the signature.

There are two sources of scans for sets from the Hamby study available
to us: scans of Hamby 44 and Hamby 252 are available from the NIST
database \citep{nist}. The physical Hamby 44 set has also been made
available to us and has been scanned locally for CSAFE at the Roy
J.~Carver High Resolution Microscopy Facility using a Sensofar confocal
light microscope. Scans in the NIST database are made with a NanoFocus
at 20x magnification. The resolutions of the two instruments are
different: the NIST scans are taken at a resolution of 1.5625 \(\mu m\)
per pixel, while the CSAFE scans are available at a resolution of 0.645
\(\mu m\) per pixel. The length of an average bullet land from Hamby (9
mm Ruger P85) is about 2 millimeter, resulting in signatures of about
1200 pixels for NIST scans, and about 3000 pixels for CSAFE scans.

In comparison, scans from the profilometer used by
\citet{chumbley, hadler} were taken at a resolution of about 0.73
\(\mu m\) per pixel. The screw driver toolmarks are about 7 mm in length
\citep{manytoolmarks1}, for a total of over 9000 pixels for the width of
these scans.

This severe limitation in the amount of available data poses the main
challenge in adapting the Chumbley score to matching bullet lands,
because of the resulting loss in power.

\hypertarget{the-chumbley-score-test}{%
\subsection*{\texorpdfstring{\emph{The Chumbley Score
Test}}{The Chumbley Score Test}}\label{the-chumbley-score-test}}
\addcontentsline{toc}{subsection}{\emph{The Chumbley Score Test}}

A digitized toolmark forms a spatial process \(z(t)\) with location
indexed by \(t\). \(t\) here denotes equally spaced pixel locations for
the striation marks under consideration. For a toolmark consisting of
\(t\) pixels, \(t = 1, ..., T\). Let further \(z^s(t)\) denote a vector
of markings of length \(s\) starting in location \(t\).

The Chumbley score algorithm takes input in form of two digitized
toolmarks:

Let \(x(t_1)\), \(t_1 = 1,2,...T_1\) and \(y(t_2)\), \(t_2 = 1,2...T_2\)
be two digitized toolmarks (where \(T_1\) and \(T_2\), the lengths of
the two marks, are not necessarily equal). The toolmarks under
consideration are potentially from two different-sources or the
same-source.

\begin{figure}

{\centering \includegraphics[width=\textwidth]{figures/sigs-profiles-1}

}

\caption{Bullet land profile (left) and the corresponding signature (right) for one of the lands of Hamby-44.}\label{fig:sigs-profiles}
\end{figure}

In a pre-processing step the two markings are smoothed using a LOWESS
\citep{lowess} with coarseness parameter \(c\). Originally, this
smoothing is intended to remove drift and (sub)class characteristics
from individual markings, however, in the setting of matching bullet
striae, we can also make use of this mechanism to separate bullet
curvature in profiles from signatures before matching signatures.
\autoref{fig:sigs-profiles} shows an example of a bullet land profile
(left) and the corresponding signature (right).

After removing sub-class structure, the Chumbley score is calculated in
two steps: an optimization step and a validation step. In the
optimization step, the two markings are aligned horizontally such that
within a pre-defined window of length \(w_o\) the correlation between
\(x(t_1)\) and \(y(t_2)\) is maximized: \[
\left(t_1^o, t_2^o\right) = \mathop{\arg \max}\limits_{1 \le t_1 \le T_1-w_o, 1 \le t_2 \le T_2-w_o} \text{cor} \left(x^{w_o} (t_1), y^{w_o}(t_2) \right)
\] This results in an optimal vertical (in-phase) shift of
\(t_1^o - t_2^o\) for aligning the two markings. We will denote the
relative optimal locations as \(t_1^*\) and \(t_2^*\), where
\(t_k^* = t_k^o/(T_k-w_o)\) for \(k=1,2\), such that
\(t_1^*, t_2^* \in [0,1]\). Once (sub-)class characteristics are
removed, the relative optimal locations should be distributed according
to a uniform distribution in \([0,1]\).

In the validation step, two sets of windows of size \(w_v\) are chosen
from both markings, see \autoref{fig:win-comparison}. In the first set,
pairs of windows are extracted from the two markings using the optimal
vertical shift as determined in the first step, whereas for the second
set the windows are extracted using a different (out-of-phase) shift.

More precisely, let us define starting points \(s_i^{(k)}\) for each
signature \(k = 1, 2\) as \begin{eqnarray}\label{eq.start}
s^{(k)}_i = 
\begin{cases}
t_k^* + i w_v & \text{ for } i < 0 \\
t_k^* + w_ o + i w_v & \text{ for } i \ge 0,
\end{cases}
\end{eqnarray} for integer values of \(i\) with
\(0 < s^{(k)}_i \le T_k - w_v\).

Same-shift pairs of length \(w_v\) are defined in \citet{hadler} as all
pairs \((s_i^{(1)}, s_i^{(2)})\) with integer values \(i\) for which
both \(s_i^{(1)}\) and \(s_i^{(2)}\) are defined. Similarly,
different-shift pairs are defined as \((s_i^{(1)}, s_{-i-1}^{(2)})\) for
all \(i\) where both \(s_i^{(1)}\) and \(s_{-i-1}^{(2)}\) are defined
(see \autoref{sketch-same-diff}).

\begin{figure}[hbtp]
\centering
\includegraphics[width=.7\textwidth]{images/sketch-same.png}

\includegraphics[width=.7\textwidth]{images/sketch-diff.png}
\caption{\label{sketch-same-diff}Sketch of same-shift pairings  (top) and different-shift pairings (bottom). Filled in rectangles show pairings resulting in correlations, unfilled rectangles are segments without a match.}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=\textwidth]{figures/win-comparison-1}

}

\caption{Two same-source markings. For convenience, the markings are moved into phase on the left and out-of phase on the right. In-phase (left) and out-of-phase (right) samples are shown by the light grey background. The Chumbley-score is based on a Mann-Whitney U test of the correlations derived from these two sets of samples.}\label{fig:win-comparison}
\end{figure}

For both same- and different-shift pairs correlations between the
markings are calculated. The intuition here is that for two markings
from the same-source the correlation for the in-phase sample should be
high, while the correlations of the out-of-phase sample provide a
measure for the base-level correlation for non-matching marks of a given
length \(w_v\). More specifically, the null hypothesis of the Chumbley
score test is stated as \textbf{\(H_o\): the markings come from two
different sources} with the alternative given as \textbf{\(H_a\): the
markings come from the same source}. A \(p\)-value for the Chumbley
score test is then computed as a Mann Whitney U statistic to compare
between in-phase sample and out-of-phase sample. A low \(p\)-value is
interpreted as a rejection of the null hypothesis in favor of the
alternative hypothesis of \emph{same-source}.

In the original method proposed in \citet{chumbley} both in-phase and
out-of-phase sample are extracted randomly, whereas \citet{hadler}
proposed the above specified deterministic rules for both samples to
make the resulting score deterministic while simultaneously avoiding
overlaps within selected marks to ensure independence.

\hypertarget{a-problem-with-failed-tests}{%
\subsection*{\texorpdfstring{\emph{A problem with failed
tests}}{A problem with failed tests}}\label{a-problem-with-failed-tests}}
\addcontentsline{toc}{subsection}{\emph{A problem with failed tests}}

Looking closer at \autoref{eq.start}, we see that by definition, some
number of tests will fail to produce a result.
{\textcolor{orange}{Note that this problem is different from erroneous test results.}}
The problem of failed tests is first mentioned in \citet{afte-chumbley}.
Unfortunately, the authors do not provide any percentage of how many
tests failed for their data. The algorithm fails to produce for two
reasons: either the number of eligible same-shift pairs is zero, or the
number of different-shift pairs is zero. Section
\ref{appendix:appxfailed} in the Appendix discusses scenarios of failed
tests in more detail.

The number of same-shift pairs will be zero, if the optimal locations
\(t_1^{o}\) and \(t_2^{o}\) are so far apart, that no segments of size
\(w_v\) are left on the same sides of the optimal locations, i.e.
\(t_1^{o} < w_v\) and \(t_2^{o} > T_2-w_o-w_v\) or
\(t_1^{o} < T_1- w_o - w_v\) and \(t_2^{o} < w_v\), i.e.~we have a
failure rate of \[
P\left( t_1^{o} < w_v \ \cap \ t_2^{o} > T_2-w_o-w_v\right) + P\left( t_1^{o} < T_1- w_o - w_v \ \cap \ t_2^{o} < w_v\right).
\] While we can assume that once (sub-)class characteristics are
removed, optimal locations \(t_i^{o}\) are uniformly distributed across
the length of the profile, we cannot assume that \(t_1^o\) and \(t_2^o\)
are independent of each other. In particular, for same-source profiles,
we would expect a strong dependency between these locations, in which
case a large difference between locations is unlikely. However, for
different-source matches, we can assume that locations are independent.
In that case, we expect a test to fail with a probability of
\(\frac{2 w_v^2}{(T_1-w_o)(T_2-w_o)}\). For an average length of \(T_i\)
of 1200 pixels, \(w_o = 120\) pixels and \(w_v = 30\) pixels this
probability is about 0.0015.

The number of possible different-shift pairs also depends on the
location of the optimal locations \(t_1^o\) and \(t_2^o\). Whenever the
optimal locations are close to the boundaries, the number of possible
pairings decreases and reaches zero, if \(t_i^{(o)} < w_v\) or
\(t_i^{(o)} > T_i-w_o- w_v\). Assuming a correlation between optimal
locations \(t_1^o\) and \(t_2^o\) of close to one for same-source
profiles, this results in an expected rate of failure of
\(2 w_v / (T_i-w_o)\), or about 5.6\% for an average length of \(T_i\)
of 1200 pixels, \(w_o = 120\) pixels and \(w_v = 30\) pixels. Assuming
independence in the optimal locations for different-source profiles the
expected probability for a failed test is, again,
\(\frac{2 w_v^2}{(T_1-w_o)(T_2-w_o)}\).

\hypertarget{a-modified-approach}{%
\subsection*{\texorpdfstring{\emph{A modified
approach}}{A modified approach}}\label{a-modified-approach}}
\addcontentsline{toc}{subsection}{\emph{A modified approach}}

While failures due to missing correlations from same-shift pairs are
unavoidable by definition of the Chumbley score, failures due to missing
correlations from different-shift pairs can be prevented by using a
different strategy in assigning pairs.

Using the same notation as in \autoref{eq.start}, we define same-shift
pairs identical to \citet{hadler} as pairs \((s_i^{(1)}, s_i^{(2)})\)
for all \(i\) where the boundary conditions of both sequences are met
simultaneously. Let us assume that this results in \(I\) pairs. Define
\(s_{(j)}^{(k)}\) to be the \(j\)th starting location in sequence
\(k = 1, 2\),
i.e.~\(s_{(1)}^{(k)} < s_{(2)}^{(k)} < ... < s_{(I)}^{(k)}\).

\begin{figure}[hbtp]
\centering
\includegraphics[width=.7\textwidth]{images/sketch-diff-2.png}
\caption{\label{sketch-diff-2}Sketch of adjusted different-shift pairings. Pairings are side from opposite ends of the two markings. For an odd number of pairings the middle pair is left out. At most one of the same-shift pairings can not be matched with a different-shift pair under algorithm (CS2). }
\end{figure}

We then define the pairs for different-shifts by matching up windows
from opposite ends of the markings, i.e.~the first pair consists of a
matchup of the first window on the first marking and the last window on
the second marking, the second pair consists of the second window on the
first marking and the second to last pair on the second marking, and so
on. In case of an odd number of pairs we need to be careful to exclude
the middle pair from this assignment: the middle pair is already part of
the same-shift pair, therefore we cannot re-use the same pair as part of
the different-shift pairs.

Mathematically, this assignment of pairs is written as:
\begin{equation}\label{eq.diff2}
\left(s_{(j)}^{(1)}, s_{(I-j+1)}^{(1)} \right) \text{ for } j = 
\begin{cases}
1, ..., I & \text{ for even } I \\
1, ..., (I-1)/2, (I-1)/2 + 2, ..., I & \text{ for odd } I
\end{cases}.
\end{equation} Note that for an odd number of same-shift correlations,
we skip the middle pair for the different-shift correlations (see also
\autoref{sketch-diff-2}). This pairing ensures that the number of
different-shift pairings is the same or at most one less than the number
of same-shift pairings in all tests. In the remainder of the paper, we
will refer to the algorithm defined by \citet{hadler} as \textbf{(CS1)}
and the suggested modified algorithm as \textbf{(CS2)} and compare their
performance on the available scans of the Hamby study.

Performance of tests is measured with respect to the errors a test makes
in situations where ground truth is known. We distinguish between two
error rates: (1) false positive and (2) false negative rate. \emph{False
positives} (or \emph{false identifications}) are situations where the
test indicates a match (i.e.~the test falsely rejects) but the markings
come from different sources. This is also known as a \emph{Type I}
error. \emph{False negatives} (or \emph{missed identifications}) are
situations where the test fails to reject, i.e.~the test indicates that
the markings come from different sources, but in fact the markings come
from the same source. This is a \emph{Type II} error. In both cases
error rates are calculated as the ratio of the number of errors observed
and the number of tests executed.

Note that in all of the following land-to-land comparisons only lands
are compared that are suitable for a comparison, i.e.~a signature can be
extracted from the scan. In particular, lands which exhibited tank rash
were removed from comparison \citep{aoas}.

\hypertarget{testing-setup}{%
\subsection*{\texorpdfstring{\emph{Testing
setup}}{Testing setup}}\label{testing-setup}}
\addcontentsline{toc}{subsection}{\emph{Testing setup}}

\hypertarget{the-data}{%
\subsubsection*{The Data}\label{the-data}}
\addcontentsline{toc}{subsubsection}{The Data}

Lands for all Hamby-44 and Hamby-252 scans are made available through
the NIST ballistics database \citep{nist} and are considered, here. Both
of these sets of scans are part of the larger Hamby study \citep{hamby}.
Each set consists of twenty known bullets (two each from ten
consecutively rifled Ruger P85 barrels) and fifteen questioned bullets
(each matching one of the ten barrels). Ground truth for both of these
Hamby sets is known and was used to assess correctness of the tests
results.

Profiles for each bullet land were extracted from scans close to the
heel of the bullet while avoiding break-off as described in
\citet{aoas}.

\hypertarget{setup}{%
\subsubsection*{Setup}\label{setup}}
\addcontentsline{toc}{subsubsection}{Setup}

Both algorithms (CS1) and (CS2) are implemented in R \citep{R}. (CS1) is
available from package \texttt{toolmaRk} \citep{toolmark}, (CS2) is
available from a modified version of the \texttt{toolmaRk} package
available from GitHub (\url{https://github.com/heike/toolmaRk}). We
applied both methods to all pairwise land-to-land comparisons of the
Hamby scans provided by NIST for a total of 85,491 land-to-land
comparisons.

\hypertarget{results}{%
\section*{Results}\label{results}}
\addcontentsline{toc}{section}{Results}

\hypertarget{failed-tests}{%
\subsection*{\texorpdfstring{\emph{Failed
Tests}}{Failed Tests}}\label{failed-tests}}
\addcontentsline{toc}{subsection}{\emph{Failed Tests}}

As described above, the Chumbley-score is based on three parameters:
coarseness \(c\) and the sizes of the optimization window \(w_o\) and
validation window \(w_v\). In a first run of results, we applied default
settings for the parameters, as suggested in \citet{hadler}:
\(w_o = 120\) pixels or about 190 \(\mu m\) (ten percent of the average
length of profiles) and coarseness \(c = 0.25\), and varied the size of
the validation window \(w_v\) in steps of 10 from 10 pixels to 60
pixels. Based on a significance level \(\alpha\) of 0.05 for the test,
this results in a correct identification of same-source and
different-source toolmarks of 93.5\% to 94.1\%, corresponding to a rate
of false negatives between 0.28 and 0.36 and a rate of false positives
between 0.05 and 0.06. However, the most prominent result we
encountered, are the high number of failed tests, i.e.~the number of
instances, in which CS1 did not return any result. \autoref{fig:fails}
shows the percentage of failed tests among the 85,491 land-to-land
comparisons of the NIST data for different values of the validation
window size \(w_v\). For same-source lands up to 12.5 percent of the
tests fail using CS1. The highest percentage of failed tests under CS2
is 1.3\% for different-source tests using a validation window size
\(w_v\) of 60 pixels. Rates of expected failures are based on simulation
runs using covariances between locations of same-source profiles of
0.854, and 0.120 for locations from different-source profiles, matching
observed covariances for the Hamby scans. Observed failure rates are
higher than expected. This might be due to remaining sub-class structure
at a coarseness of 0.25 resulting in a distribution of optimal locations
different from the assumed uniform.

\begin{figure}

{\centering \includegraphics[width=0.7\textwidth]{figures/fails-1}

}

\caption{Percent of failed land-to-land comparisons using an optimization window $w_o = 120$ and a coarseness of $c = 0.25$. With an increase in the size of the validation window  a higher percentage of tests fails under both methods (CS1) and (CS2), but the percentage of failed tests is much smaller under (CS2). Observed failure rates of (CS1) are higher than expected rates.}\label{fig:fails}
\end{figure}

\hypertarget{coarseness}{%
\subsection*{\texorpdfstring{\emph{Coarseness}}{Coarseness}}\label{coarseness}}
\addcontentsline{toc}{subsection}{\emph{Coarseness}}

The purpose of the coarseness parameter is to remove (sub-)class
characteristics from profiles before comparisons for matching.
\citet{hadler} suggest a coarseness parameter of 0.25 in the setting of
toolmark comparisons. For bullet lands, coarseness might need to be
adjusted because of the strong effect bullet curvature has on profiles.

\autoref{fig:profile-sketch} gives an overview of the effect of
different coarseness parameters: from left to right, coarseness levels
\(c\) are varied in steps of 0.05 from 0.1 to 0.3. The top row shows
resulting signatures after smoothing the profile shown in
\autoref{fig:sigs-profiles} with different levels of coarseness. The
histograms in the bottom row show the relative optimal location \(t^*\).
Optimal locations are distributed uniformly once (sub-)class
characteristics are removed. However, for coarseness values of
\(c > 0.20\) we see quite distinct boundary effects: optimal locations
\(t^*\) are found at the very extreme ends of a profile more often than
one would expect based on a uniform distribution.

The key effect of the optimal locations and thereby the coarseness is
seen in the number of failed tests. Irrespective of whether CS1 or CS2
is being used, if the relative optimal locations are at the boundaries
we will see an increase in the number of failed tests. A balance is
therefore needed in the selection of the coarseness parameter which
reduces the boundary effect but does not remove important individual
characteristics. Based on \autoref{fig:profile-sketch} a coarseness
value of \(c = 0.15\) seems to be best suited to strike this balance for
this example. For the remainder of the analysis, we will use this value
for \(c\).

\begin{figure}

{\centering \includegraphics[width=\textwidth]{figures/profile-sketch-1}

}

\caption{Overview of the effect of different coarseness parameters $c$ on the profile shown in Figure \ref{fig:sigs-profiles} (top). The bottom row shows histograms of  the (relative) optimal locations $t^o$ identified in the optimization step for different values of the coarseness parameter $c$. }\label{fig:profile-sketch}
\end{figure}

\hypertarget{error-rate-assessment}{%
\subsection*{\texorpdfstring{\emph{Error rate
assessment}}{Error rate assessment}}\label{error-rate-assessment}}
\addcontentsline{toc}{subsection}{\emph{Error rate assessment}}

>>>>

% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{adjustbox}
\usepackage{multirow}
\usepackage{subfig}
\usepackage[table]{xcolor}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{hyperref}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}

\usepackage{url} % not crucial - just used below for the URL

@@

\if0\blind
{
  \title{\bf Adapting the Chumbley Score to match striae on Land Engraved Areas (LEAs) of bullets}

  \author{
        Ganesh Krishnan,\\ %\\
    Master of Science (M.Sc.)\\
    % \\
     and \\     Prof.~Heike Hofmann,\\ %\\
    Doctor of Philosophy (Ph.D.)\\
    % \\
      }
    \maketitle
  \bigskip
  \begin{center}
  {\textbf{Affiliation(s)}\\
  Department of Statistics, Iowa State University and CSAFE (Center for
  Statistics and Applications in Forensic Evidence)\\
  \bigskip
  \textbf{Source of funding}\\
  The authors gratefully acknowledge funding from CSAFE through
  cooperative agreement \# 70NANB15H176 between NIST and Iowa State
  University.\\
  \bigskip
  \textbf{Acknowledgement}\\
  We would like to thank the efforts of CSAFE and Roy J Carver Microscopy
  lab in scanning the Hamby-44 set and providing the scans to us. We would
  also like to thank NIST for providing the Hamby-44 bullets to us for
  scanning.\\
  \bigskip
  \textbf{Information on presentation of work}\\
  This work was presented in the form of a poster in the AFTE 2018 49th
  Annual Training Seminar, held from June 3rd to June 8th, 2018 in
  Charleston, West Virginia}
  \end{center}

@@
